Distributional Regression Forests for Probabilistic Precipitation Forecasting in Complex Terrain

### 2018-11-02 pkg/disttree/inst/draft/draft.Rnw
final Z revision of main manuscript, adapting much of the discussion further touch-ups in remaining materials
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2375 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: a394dd57fb692b04399df40ba94b573455f36c1f

The general distributional notation is exemplified in all three steps for the
Gaussian distribution left-censored at zero. The latter is employed in the empirical 
case study in Section~\ref{sec:precipitation} to model power-transformed daily 
precipitation amounts.

@@

To capture both location and scale of the probabilistic precipitation forecasts
while accounting for a point mass at zero (i.e., dry days without
rain), a Gaussian distribution left-censored at zero (for short: zero-censored Gaussian)
with location parameter $\mu$ and scale parameter $\sigma$ is employed. Therefore,
the corresponding log-likelihood function with parameter vector 
$\bm{\theta} = (\mu, \sigma)$ is
\begin{equation}
\ell(\mu, \sigma; Y) = 
\begin{cases}
    \log\left\{\frac{1}{\sigma} \cdot \phi\left(\frac{Y - \mu}{\sigma}\right) \right\}, & \text{if } Y > 0\\[0.2cm]
    \log\left\{\Phi\left(\frac{-\mu}{\sigma}\right)\right\}, & \text{if } Y = 0
\end{cases}
\end{equation}

where $\phi$ and $\Phi$ are the probability density function and the 
distribution function of the standard normal distribution $\mathcal{N}(0,1)$. 
Other distributions $\mathcal{D}$ and corresponding log-likelihoods
$\ell(\mu, \sigma; Y)$ could be set up in the same way, e.g., for
censored shifted gamma distributions \citep{Scheuerer+Hamill:2015} or
zero-censored logistic distributions \citep{Gebetsberger+Messner+Mayr:2017}.

@@

Typically, a single global model $\mathcal{D}(Y, \bm{\theta})$ is not sufficient
for reasonably representing the response distribution and covariates
$\bold{Z} = Z_1, \dots, Z_m \in \mathcal{Z}$ are
employed to capture differences in the distribution parameters $\bm{\theta}$.
In weather forecasting, these covariates typically include the output
from numerical weather prediction systems and/or lagged meteorological observations.

@@

Regression trees generally take a different approach for automatically
including covariates in a data-driven way and allowing for abrupt
changes, nonlinear and non-additive effects, and interactions. In the context of
distributional models the goal is to partition the covariate space
$\mathcal{Z}$ recursively into disjoint segments so that a homogenous
distributional model for the response $Y$ can be found with segment-specific
parameters. More specifically, the $B$ disjoint segments $\mathcal{B}_b$
($b = 1, \dots, B$) partition the covariate space
\begin{equation}
\label{eq:partition}
\mathcal{Z} = \dot{\bigcup\limits_{b = 1, \ldots, \textit{B}}} \mathcal{B}_b,
\end{equation}
and a local distributional model $\mathcal{D}(Y, \bm{\theta}^{(b)})$
(i.e., with segment-specific parameters $\bm{\theta}^{(b)}$) is fitted to the
response $Y$ in each segment.

@@

To sum up, distributional trees are fitted recursively via:
\begin{enumerate}
\item Estimate $\bm{\hat \theta}$ via maximum likelihood for the observations
  in the current subsample.
\item Test for associations or instabilities of the scores $s(\bm{\hat \theta}, y_i)$ 
  and $Z_{l,i}$ for each partitioning variable~$Z_l$ ($l = 1, \dots, m$).
\item Split the sample along the partitioning variable $Z_l^*$ with the
  strongest association or instability. Choose the breakpoint with the highest
  improvement in the log-likelihood or the highest discrepancy.
\item Repeat steps 1--3 recursively in the subsamples until these become too
  small or there is no significant association/instability (or some other
  stopping criterion is reached).
\end{enumerate}
%
Different inference techniques can be used for assessing the association between
scores and covariates in step~3. In the following we use the general class of
permutation tests introduced by \cite{Hothorn+Hornik+VanDeWiel:2006} which is
also the basis of conditional inference trees \citep[CTree,][]{Hothorn+Hornik+Zeileis:2006}.
Alternatively, one could use asymptotic M-fluctuation tests for parameter
instability \citep{Zeileis+Hornik:2007} as in model-based recursive partitioning
\citep[MOB,][]{Zeileis+Hothorn+Hornik:2008}. More details are provided in
Appendix~\ref{app:tree}.

For obtaining probabilistic predictions from the tree for a (possibly new) set
of covariates $\bold{z} = (z_1, \ldots, z_m)$, the observation simply has to
be ``sent down'' the tree and the corresponding segment-specific MLE has to be
obtained. This can also be understood as a weighted MLE where the weights
select those observations from the learning sample that fall into the same
segment:
\begin{equation}
w^{\text{tree}}_i(\bold{z}) = \sum_{b=1}^B \mathbf{1}((\bold{z}_i \in \mathcal{B}_b) \land (\bold{z} \in \mathcal{B}_b)),
\end{equation}
where $\mathbf{1}(\cdot)$ is the indicator function. The predicted distribution
for a given $\bold{z}$ is then fully specified by the estimated parameter
$\bm{\hat \theta}(\bold{z})$ where
\begin{equation}
\bm{\hat \theta}(\bold{z}) = \argmax_{\bm{\theta} \in \bm{\Theta}} \sum_{i=1}^n w^{\text{tree}}_i(\bold{z}) \cdot \ell(\bm{\theta}; y_i).
\end{equation}
\hlc{Even though this parameter depends on the given $\bold{z}$ it has to be noted that
it is the same for all other observations that are in the same terminal node of 
the tree as $\bold{z}$ and therefore does not necessarily have to be recalculated for
each of them.}

@@

\hlc{The idea of random forests is to learn an ensemble of trees, each on a different
learning data obtained through resampling (bootstrap or subsampling). In each node 
only a random subset of the covariates $\bold{Z}$ is considered to reduce the 
correlation among the trees and to stabilize the variance of the model.
For a simple regression random forest the mean of predictions over
all trees is considered. In that way changes in the location  of the response 
across the covariates are detected} (e.g., in Breiman and Cutler's random forests, 
\citealp{Breiman:2001}).
\hlc{This idea is now taken one step further by embedding it in a distributional 
framework based on loglikelihood estimation. \emph{Distributional forests} 
employ an ensemble of $T$~\emph{distributional trees} which pick up changes in 
the ``direction'' of any distribution parameter by considering the full score 
vector for choosing splitting variables and split points.}

@@

Distributional forests share some concepts and algorithmic aspects 
with other generalizations of Breiman and Cutler's random forests. Nearest 
neighbor weights are employed for aggregation in survival forests 
\citep{Hothorn+Lausen+Benner:2004},
quantile regression forests \citep{Meinshausen:2006},
transformation forests \citep{Hothorn+Zeileis:2017},
and generalized random forests for causal inferences 
\citep{Athey+Tibshirani+Wager:2017}.
These procedures aggregate over trees fitted to specific score functions 
(e.g., log rank scores in survival trees, model residuals in transformation 
or generalized forests). Distributional forests, in contrast to these 
nonparametric approaches, provide a compromise between model flexibility 
and interpretability: The parameters of a problem-specific distribution 
(zero-censored Gaussian for precipitation) have a clear meaning but 
may depend on external variables in a quite general way.

@@

Many statistical weather forecasting models leverage the strengths of modern
numerical ensemble prediction systems (EPSs; see \citealp{Bauer+Thorpe+Brunet:2015}).
\hlc{EPSs not only predict the most likely future state of the atmosphere but provide
information about the uncertainty for a specific quantity and weather situation.
This is done by running the NWP model several times using slightly perturbed 
initial conditions and model specifications to account for uncertainties in both,
the initial state and the model and its parametrizations.
One frequently used method based on distributional regression models is the 
ensemble model output statistics (EMOS) approach first proposed by}
\cite{Gneiting+Raftery+Westveld:2005} \hlc{to produce high-quality forecasts for
specific quantities and sites.  
In case of precipitation forecasting EMOS typically uses the ensemble mean of 
``total precipitation''} (\emph{tp}) \hlc{forecasts as predictor for the location 
parameter $\mu$ and the corresponding ensemble standard deviation for the 
scale parameter $\sigma$ of the statistical model based on the assumption
of the response following a Gaussian distribution left-censored at zero.
This distributional approach of modeling both parameters allows to correct 
for possible errors of the NWP ensemble in both, the expectation but also the 
uncertainty of a specific forecast. In order to assure positivity of the 
scale parameter a log-link is applied. Hence, the two distribution parameters 
are modeled by the linear functions} 
$\mu = \beta_0 + \beta_1 \cdot \emph{mean}(\emph{tp})$ and  
$\log(\sigma) = \gamma_0 + \gamma_1 \cdot \emph{sd}(\emph{tp})$
with regression coefficients $\beta_0$, $\beta_1$, $\gamma_0$ and $\gamma_1$,  
\hlc{hereafter referred to as a basic EMOS model} \citep{Gebetsberger+Messner+Mayr:2017}.

% FIXME: notation with or without power transformation parameter?

% with:
%$\mu = \beta_0 + \beta_1 \cdot \mbox{mean}(\mbox{tp}^{\frac{1}{p}})$ and  
%$\log(\sigma) = \gamma_0 + \gamma_1 \cdot \mbox{sd}(\mbox{tp}^{\frac{1}{p}})$
%with power transformation parameter $\frac{1}{p}$ and 
%regression coefficients $\beta_0$, $\beta_1$, $\gamma_0$ and $\gamma_1$,  
%\hlc{hereafter referred to as a basic EMOS model \citep{Gebetsberger+Messner+Mayr:2017}.}

While this approach alone is already highly effective in the plains,
it typically does not perform as well in complex terrain due to unresolved
effects in the \hlc{NWP system} \citep{Bauer+Thorpe+Brunet:2015}. 
For example, in the Tyrolean Alps -- considered
in the following case study -- the NWP grid cells of $50 \times 50$
km$^2$ are too coarse to capture single mountains, narrow valleys, etc. 
Therefore, it is often possible to substantially improve the predictive 
performance of the EMOS by including additional predictor variables, either 
from local meteorological observations or an NWP model. Unfortunately, it is 
typically unknown which variables are relevant for improving the predictions. 
Simply including all available variables may be computationally burdensome and
can lead to overfitting but, on the other hand, excluding too many variables
may result in a loss of valuable information. Therefore, selecting
the relevant variables and interactions among all possible covariates is 
crucial for improving the statistical forecasting model.

In the following, it is illustrated how distributional forests can
solve this problem without requiring prior expert knowledge.  For fitting the
forest only the response distribution and the list of potential predictor
variables need to be specified (along with a few algorithmic details) and then
the relevant variables, interactions, and potentially nonlinear effects are
determined automatically in a data-driven way. Here, we employ a zero-censored
Gaussian distribution and 80~predictor variables computed from ensemble means
and spreads of various NWP outputs. The predictive performance of the forest is
compared to three other zero-censored Gaussian models: (a)~a basic
EMOS, (b)~a GAMLSS with prespecified effects and interactions based on
meteorological knowledge/experience, and (c)~a boosted GAMLSS with automatic
selection of smooth additive terms based on all 80~predictor variables.

@@

\begin{table}[t!]
%\resizebox{1.4\textwidth}{!}{
\begin{minipage}{\textwidth}
\caption[Table caption text]{Basic covariates together with the number ({\#}) and the type of variations.
Time periods indicate aggregation time periods in hours after NWP model
initialization (e.g., 6--30 corresponds to +6\,h to +30\,h ahead forecasts,
0600\,UTC to 0600\,UTC of the next day). $^*$Minimum values of \emph{dswrf}
over 24\,h are always zero and thus neglected.}
\label{tab:covariates}
\begin{tabular}{ l  c  l }
\hline
Basic covariates & {\#} & Variations\\
\hline
\emph{tp}: total precipitation,           & 12  & ensemble mean of sums over 24h, \\
\hspace*{0.5cm} power transformed (by $1.6^{-1}$) &    & ensemble std. deviation of sums over 24h, \\ 
\emph{cape}: convective available     &    & ensemble minimum of sums over 24h, \\
\hspace*{0.9cm} potential energy, &    & ensemble maximum of sums over 24h\\
\hspace*{0.9cm} power transformed (by $1.6^{-1}$)&    & \qquad all for 6--30 \\
\hspace*{\fill}                            &    & ensemble mean of sums over 6h\\
\hspace*{\fill}                            &    & \qquad for 6--12, 12--18, 18--24, 24--30 \\
\hspace*{\fill}                            &    & ensemble std. deviation of sums over 6h\\
\hspace*{\fill}                            &    & \qquad for 6--12, 12--18, 18--24, 24--30 \\
\hline
\emph{dswrf}: downwards short wave      & 6 & ensemble mean of mean values, \\
\hspace*{1.13cm} radiation flux (``sunshine'') &   & ensemble mean of minimum values$^*$,\\
\emph{msl}: mean sea level pressure     &   &  ensemble mean of maximal values,\\
\emph{pwat}: precipitable water         &   & ensemble std. deviation of mean values,\\
\emph{tmax}: 2m maximum temperature     &   & ensemble std. deviation of minimum values$^*$,\\
\hspace*{1.15cm}                          &   & ensemble std. deviation of maximal values,\\
\emph{tcolc}: total column-integrated   &   & \qquad all over 6--30 \\
\hspace*{0.95cm} condensate               &   & \\
\emph{t500}: temperature on 500 hPa     &   & \\
\hspace*{1cm}                             &   & \\
\emph{t700}: temperature on 700 hPa     &   & \\
\hspace*{1cm}                             &   & \\
\emph{t850}: temperature on 850 hPa     &   & \\
\hspace{1cm}                              &   & \\
\hline
\emph{tdiff500850}: temperature         & 3 & ensemble mean of difference in mean,\\
\hspace*{\fill} difference 500 to 850 hPa &   & ensemble minimum of difference in mean,\\
\emph{tdiff500700}: temperature         &   & ensemble maximum of difference in mean\\
\hspace*{\fill} difference 500 to 700 hPa &   & \qquad all over 6--30 \\
\emph{tdiff700850}: temperature         &   & \\
\hspace*{\fill} difference 700 to 850 hPa &   & \\
\hline
\emph{msl{\_}diff}: mean sea level pressure & 1 & \emph{msl{\_}mean{\_}max} $-$ \emph{msl{\_}mean{\_}min}\\
\hspace{1.6cm} difference            &   & \qquad over 6--30 \\
\hline
\end{tabular}
\end{minipage} 
%}
\end{table}

@@

Observations are obtained for 95~stations all over Tyrol and surroundings,
providing 24-hour precipitation sums measured at 0600\,UTC and rigorously
quality-checked by the National Hydrographical Service. NWP outputs
are obtained from the second generation reforecast data set of the
global ensemble forecast system \citep[GEFS,][]{Hamill+Bates+Whitaker:2013}. 
This data set consists of an 11-member ensemble based on a fixed version of the 
numerical model and a horizontal grid spacing of about {$50 \times 50$ km$^2$}
initialized daily at 0000~UTC from December 1984 to present 
providing forecasts on a 6-hourly temporal resolution. Each of the 
11 ensemble members uses slightly different perturbed initial 
conditions to predict the situation-specific uncertainty 
of the atmospheric state.

@@

To remove large parts of the skewness of precipitation data, 
a power transformation \citep{Box+Cox:1964} is often applied, e.g., using
cubic \citep{Stidd:1973} or square root \citep{Hutchinson:1998} transformations.
However, the power parameter may vary for different climatic zones or temporal 
aggregation periods and hence we follow \cite{Stauffer+Mayr+Messner:2017} 
in their choice of $1.6^{-1}$ as a suitable power parameter for the
region of Tyrol. The same power transformation is applied to both the
observed precipitation sums and the NWP outputs ``total precipitation'' (\emph{tp}) and
``convective available potential energy'' (\emph{cape}).

@@

\item \emph{Distributional forest:} All 80~predictor variables are
  considered for learning a forest of 100 trees. Subsampling is employed 
  for each tree using a third of the predictors
  in each split of the tree (\code{mtry}). Parameters are estimated by
  adaptive local likelihood based on the forest weights as described
  in Section~\ref{sec:methods}. 
  \hlc{As stopping criteria the minimal number of observations to perform a split} 
  (\code{minsplit}) \hlc{is set to $50$, the minimal number of observations in a 
  terminal node} (\code{minbucket}) \hlc{is set to $20$ and the significance level 
  for variable selection} (\code{alpha}) \hlc{is kept at the default value $1$ such 
  that no early-stopping (sometimes also referred to as ``pre-pruning'') is applied
  based on results of the statistical tests.}
  % stopping criteria:
  % forest_minsplit = 50, forest_minbucket = 20, forest_mincrit = 0

@@

In a first step, we show a detailed comparison of the competing models
for one observation site, Axams in Tyrol (in the center of the study area,
see Figure~\ref{fig:all-map}). As for all other stations, daily precipitation observations and
numerical weather predictions are available for the month of July from 1985
through 2012. In Figure~\ref{fig:axams-dist} in the introduction the probabilistic
forecasts from the distributional forest, trained on 1985--2008, for July 24
in 2009--2012 have already been shown as a motivational example. The figure
shows that the model properly
depicts the point masses at zero (i.e., the probability of a dry day) and
the forecasted probability density function for the total amount of precipitation.
The four sample forecasts differ considerably in location $\mu$, scale $\sigma$, and the amount of
censoring while conforming quite well with the actual observations
from these days. While this is a nice illustrative example we are  
interested in the overall predictive performance and calibration of the
distributional fits.

To assess calibration Figure~\ref{fig:axams-qq} shows residual QQ~plots for
out-of-sample predictions (2009--2012) from the different models trained on
1985--2008. Due to the point masses at zero 100 draws from the randomized
quantile residuals \citep{Dunn+Smyth:1996} are plotted in semi-transparent
gray. Overall, the randomized quantile residuals conform quite well with the
theoretical standard normal quantile (i.e., form a straight line close to the 
diagonal), indicating that all four models are sufficiently well calibrated. 
\hlc{This is also supported by the corresponding probability integral transform}
(PIT, \citealp{Gneiting+Balabdaoui+Raftery:2007}) \hlc{histograms for station
Axams in Supplement B} \citep{Schlosser+Hothorn+Stauffer:2018b} \hlc{which 
contains a more detailed explanation of residual QQ plots and PIT histograms.}

>>>>

The general distributional notation is exemplified in all three steps using the
Gaussian distribution left-censored at zero (for short: zero-censored Gaussian).
The latter is employed in the empirical case study in Section~\ref{sec:precipitation}
to model power-transformed daily precipitation amounts.

@@

To capture both location and scale of the probabilistic precipitation forecasts
while accounting for a point mass at zero (i.e., dry days without
rain), a zero-censored Gaussian distribution
with location parameter $\mu$ and scale parameter $\sigma$ is employed. Therefore,
the corresponding log-likelihood function with parameter vector 
$\bm{\theta} = (\mu, \sigma)$ is
\begin{equation}
\ell(\mu, \sigma; Y) = 
\begin{cases}
    \log\left\{\frac{1}{\sigma} \cdot \phi\left(\frac{Y - \mu}{\sigma}\right) \right\}, & \text{if } Y > 0\\[0.2cm]
    \log\left\{\Phi\left(\frac{-\mu}{\sigma}\right)\right\}, & \text{if } Y = 0
\end{cases}
\end{equation}

where $\phi$ and $\Phi$ are the probability density function and the cumulative
distribution function of the standard normal distribution $\mathcal{N}(0,1)$. 
Other distributions $\mathcal{D}$ and corresponding log-likelihoods
$\ell(\mu, \sigma; Y)$ could be set up in the same way, e.g., for
censored shifted gamma distributions \citep{Scheuerer+Hamill:2015} or
zero-censored logistic distributions \citep{Gebetsberger+Messner+Mayr:2017}.

@@

Typically, a single global model $\mathcal{D}(Y, \bm{\theta})$ is not sufficient
for reasonably representing the response distribution. Therefore, covariates
$\bold{Z} = Z_1, \dots, Z_m \in \mathcal{Z}$ are
employed to capture differences in the distribution parameters $\bm{\theta}$.
In weather forecasting, these covariates typically include the output
from numerical weather prediction systems and/or lagged meteorological observations.

@@

Regression trees generally take a different approach for automatically
including covariates in a data-driven way and allowing for abrupt
changes, nonlinear and non-additive effects, and interactions. In the context of
distributional models the goal is to partition the covariate space
$\mathcal{Z}$ recursively into disjoint segments so that a homogenous
distributional model for the response $Y$ can be found for each segment with segment-specific
parameters. More specifically, the $B$ disjoint segments $\mathcal{B}_b$
($b = 1, \dots, B$) partition the covariate space
\begin{equation}
\label{eq:partition}
\mathcal{Z} = \dot{\bigcup\limits_{b = 1, \ldots, \textit{B}}} \mathcal{B}_b,
\end{equation}
and a local distributional model $\mathcal{D}(Y, \bm{\theta}^{(b)})$
(i.e., with segment-specific parameters $\bm{\theta}^{(b)}$) is fitted to the
response $Y$ in each segment.

@@

To sum up, distributional trees are fitted recursively via:
\begin{enumerate}
\item Estimate $\bm{\hat \theta}$ via maximum likelihood for the observations
  in the current subsample.
\item Test for associations (or instabilities) of the scores $s(\bm{\hat \theta}, y_i)$ 
  and $Z_{l,i}$ for each partitioning variable~$Z_l$ ($l = 1, \dots, m$).
\item Split the sample along the partitioning variable $Z_l^*$ with the
  strongest association or instability. Choose the breakpoint with the highest
  improvement in the log-likelihood or the highest discrepancy.
\item Repeat steps 1--3 recursively in the subsamples until these become too
  small or there is no significant association/instability (or some other
  stopping criterion is reached).
\end{enumerate}
%
Different inference techniques can be used for assessing the association between
scores and covariates in step~3. In the following we use the general class of
permutation tests introduced by \cite{Hothorn+Hornik+VanDeWiel:2006} which is
also the basis of conditional inference trees \citep[CTree,][]{Hothorn+Hornik+Zeileis:2006}.
Alternatively, one could use asymptotic M-fluctuation tests for parameter
instability \citep{Zeileis+Hornik:2007} as in model-based recursive partitioning
\citep[MOB,][]{Zeileis+Hothorn+Hornik:2008}. More details are provided in
Appendix~\ref{app:tree}.

For obtaining probabilistic predictions from the tree for a (possibly new) set
of covariates $\bold{z} = (z_1, \ldots, z_m)$, the observation simply has to
be ``sent down'' the tree and the corresponding segment-specific MLE has to be
obtained. \hlc{Thus, in practice $\bm{\hat \theta}(\bold{z})$ does not have to be recalculated
for each new $\bold{z}$ but one can simply extract the parameter estimates for
the corresponding segment which have been computed already while
learning the tree. However, to understand this estimator conceptually it is useful
to denote it as a weighted MLE} where the weights
select those observations from the learning sample that fall into the same
segment:
\begin{equation}
w^{\text{tree}}_i(\bold{z}) = \sum_{b=1}^B \mathbf{1}((\bold{z}_i \in \mathcal{B}_b) \land (\bold{z} \in \mathcal{B}_b)),
\end{equation}
where $\mathbf{1}(\cdot)$ is the indicator function. The predicted distribution
for a given $\bold{z}$ is then fully specified by the estimated parameter
$\bm{\hat \theta}(\bold{z})$ where
\begin{equation}
\bm{\hat \theta}(\bold{z}) = \argmax_{\bm{\theta} \in \bm{\Theta}} \sum_{i=1}^n w^{\text{tree}}_i(\bold{z}) \cdot \ell(\bm{\theta}; y_i).
\end{equation}

@@

\hlc{The idea of random forests is to learn an ensemble of trees, each on a different
learning data obtained through resampling (bootstrap or subsampling). In each node 
only a random subset of the covariates $\bold{Z}$ is considered for splitting to reduce the 
correlation among the trees and to stabilize the variance of the model.
For a simple regression random forest the mean of predictions over
all trees is considered. In that way changes in the location  of the response 
across the covariates are detected} (e.g., in Breiman and Cutler's random forests, 
\citealp{Breiman:2001}).
\hlc{This idea is now taken one step further by embedding it in a distributional 
framework based on maximum-likelihood estimation. \emph{Distributional forests} 
employ an ensemble of $T$~\emph{distributional trees} which pick up changes in 
the ``direction'' of any distribution parameter by considering the full likelihood
and corresponding score function for choosing splitting variables and split points.}

@@

Distributional forests share some concepts and algorithmic aspects 
with other generalizations of Breiman and Cutler's random forests. Nearest 
neighbor weights are employed for aggregation in survival forests 
\citep{Hothorn+Lausen+Benner:2004},
quantile regression forests \citep{Meinshausen:2006},
transformation forests \citep{Hothorn+Zeileis:2017},
and generalized random forests for causal inferences 
\citep{Athey+Tibshirani+Wager:2017}.
These procedures aggregate over trees fitted to specific score functions 
(e.g., log rank scores in survival trees, model residuals in transformation 
or generalized forests). Distributional forests, in contrast to 
nonparametric approaches, provide a compromise between model flexibility 
and interpretability: The parameters of a problem-specific distribution 
(zero-censored Gaussian for precipitation) have a clear meaning but 
may depend on external variables in a quite general way.

@@

Many statistical weather forecasting models leverage the strengths of modern
numerical ensemble prediction systems (EPSs; see \citealp{Bauer+Thorpe+Brunet:2015}).
\hlc{EPSs not only predict the most likely future state of the atmosphere but provide
information about the uncertainty for a specific quantity and weather situation.
This is done by running the NWP model several times using slightly perturbed 
initial conditions and model specifications to account for uncertainties in both,
the initial atmospheric state and the NWP model (and its parametrizations).
One frequently-used method based on distributional regression models is the 
ensemble model output statistics (EMOS) approach first proposed by}
\cite{Gneiting+Raftery+Westveld:2005} \hlc{to produce high-quality forecasts for
specific quantities and sites.  
In case of precipitation forecasting, EMOS typically uses the ensemble mean of 
``total precipitation'' (\emph{tp}) forecasts as the predictor for the location 
parameter $\mu$ and the corresponding ensemble standard deviation for the 
scale parameter $\sigma$, e.g., assuming the observations to follow a zero-censored
Gaussian distribution.
This distributional approach of modeling both parameters allows to correct 
for possible errors of the NWP ensemble in both, the expectation but also the 
uncertainty of a specific forecast. Thus, a basic EMOS specification typically
models the two distribution parameters by two linear predictors, e.g.,} 
$\mu = \beta_0 + \beta_1 \cdot \mathrm{mean}(\emph{tp})$ and  
$\log(\sigma) = \gamma_0 + \gamma_1 \cdot \log(\mathrm{sd}(\emph{tp}))$
with regression coefficients $\beta_0$, $\beta_1$, $\gamma_0$, and $\gamma_1$
\citep[where the log link assures positivity of the scale parameter, following][]{Gebetsberger+Messner+Mayr:2017}.

% FIXME: notation with or without power transformation parameter?
%
%$\mu = \beta_0 + \beta_1 \cdot \mbox{mean}(\mbox{tp}^{\frac{1}{p}})$ and  
%$\log(\sigma) = \gamma_0 + \gamma_1 \cdot \mbox{sd}(\mbox{tp}^{\frac{1}{p}})$
%with power transformation parameter $\frac{1}{p}$

While this approach alone is already highly effective in the plains,
it typically does not perform as well in complex terrain due to unresolved
effects in the \hlc{NWP system} \citep{Bauer+Thorpe+Brunet:2015}. 
For example, in the Tyrolean Alps -- considered
in the following case study -- the NWP grid cells of $50 \times 50$
km$^2$ are too coarse to capture single mountains, narrow valleys, etc. 
Therefore, it is often possible to substantially improve the predictive 
performance of a basic EMOS by including additional predictor variables, either 
from local meteorological observations or an NWP model. Unfortunately, it is 
typically unknown which variables are relevant for improving the predictions. 
Simply including all available variables may be computationally burdensome and
can lead to overfitting but, on the other hand, excluding too many variables
may result in a loss of valuable information. Therefore, selecting
the relevant variables and interactions among all possible covariates is 
crucial for improving the statistical forecasting model.

In the following, it is illustrated how distributional forests can
solve this problem without requiring prior expert knowledge about the meteorological covariates.
For fitting the forest only the response distribution and the list of potential predictor
variables need to be specified (along with a few algorithmic details) and then
the relevant variables, interactions, and potentially nonlinear effects are
determined automatically in a data-driven way. Here, we employ a zero-censored
Gaussian distribution and 80~predictor variables computed from ensemble means
and spreads of various NWP outputs. The predictive performance of the forest is
compared to three other zero-censored Gaussian models: (a)~a basic
EMOS, (b)~a GAMLSS with prespecified effects and interactions based on
meteorological knowledge/experience, and (c)~a boosted GAMLSS with automatic
selection of smooth additive terms based on all 80~predictor variables.

@@

\begin{table}[t!]
%\resizebox{1.4\textwidth}{!}{
\begin{minipage}{\textwidth}
\caption[Table caption text]{Basic covariates together with the number ({\#}) and the type of variations.
Time periods indicate aggregation time periods in hours after NWP model
initialization (e.g., 6--30 corresponds to +6\,h to +30\,h ahead forecasts,
0600\,UTC to 0600\,UTC of the next day). Note: $^*$Minimum values of \emph{dswrf}
over 24\,h are always zero and thus neglected.}
\label{tab:covariates}
\begin{tabular}{ l  c  l }
\hline
Basic covariates & {\#} & Variations\\
\hline
\emph{tp}: total precipitation,           & 12  & ensemble mean of sums over 24h, \\
\hspace*{0.5cm} power transformed (by $1.6^{-1}$) &    & ensemble std.\ deviation of sums over 24h, \\ 
\emph{cape}: convective available     &    & ensemble minimum of sums over 24h, \\
\hspace*{0.9cm} potential energy, &    & ensemble maximum of sums over 24h\\
\hspace*{0.9cm} power transformed (by $1.6^{-1}$)&    & \qquad all for 6--30 \\
\hspace*{\fill}                            &    & ensemble mean of sums over 6h\\
\hspace*{\fill}                            &    & \qquad for 6--12, 12--18, 18--24, 24--30 \\
\hspace*{\fill}                            &    & ensemble std.\ deviation of sums over 6h\\
\hspace*{\fill}                            &    & \qquad for 6--12, 12--18, 18--24, 24--30 \\
\hline
\emph{dswrf}: downwards short wave      & 6 & ensemble mean of mean values, \\
\hspace*{1.13cm} radiation flux (``sunshine'') &   & ensemble mean of minimum values$^*$,\\
\emph{msl}: mean sea level pressure     &   &  ensemble mean of maximal values,\\
\emph{pwat}: precipitable water         &   & ensemble std.\ deviation of mean values,\\
\emph{tmax}: 2m maximum temperature     &   & ensemble std.\ deviation of minimum values$^*$,\\
\hspace*{1.15cm}                          &   & ensemble std.\ deviation of maximal values,\\
\emph{tcolc}: total column-integrated   &   & \qquad all over 6--30 \\
\hspace*{0.95cm} condensate               &   & \\
\emph{t500}: temperature on 500 hPa     &   & \\
\hspace*{1cm}                             &   & \\
\emph{t700}: temperature on 700 hPa     &   & \\
\hspace*{1cm}                             &   & \\
\emph{t850}: temperature on 850 hPa     &   & \\
\hspace{1cm}                              &   & \\
\hline
\emph{tdiff500850}: temperature         & 3 & ensemble mean of difference in mean,\\
\hspace*{\fill} difference 500 to 850 hPa &   & ensemble minimum of difference in mean,\\
\emph{tdiff500700}: temperature         &   & ensemble maximum of difference in mean\\
\hspace*{\fill} difference 500 to 700 hPa &   & \qquad all over 6--30 \\
\emph{tdiff700850}: temperature         &   & \\
\hspace*{\fill} difference 700 to 850 hPa &   & \\
\hline
\emph{msl{\_}diff}: mean sea level pressure & 1 & \emph{msl{\_}mean{\_}max} $-$ \emph{msl{\_}mean{\_}min}\\
\hspace{1.6cm} difference            &   & \qquad over 6--30 \\
\hline
\end{tabular}
\end{minipage} 
%}
\end{table}

@@

Observations are obtained for 95~stations all over Tyrol and surroundings,
providing 24-hour precipitation sums measured at 0600\,UTC and rigorously
quality-checked by the National Hydrographical Service. NWP outputs
are obtained from the second-generation reforecast data set of the
global ensemble forecast system \citep[GEFS,][]{Hamill+Bates+Whitaker:2013}. 
This data set consists of an 11-member ensemble based on a fixed version of the 
numerical model and a horizontal grid spacing of about {$50 \times 50$ km$^2$}
initialized daily at 0000~UTC from December 1984 to present 
providing forecasts on a 6-hourly temporal resolution. Each of the 
11 ensemble members uses slightly different perturbed initial 
conditions to predict the situation-specific uncertainty 
of the atmospheric state.

@@

To remove large parts of the skewness of precipitation data, 
a power transformation \citep{Box+Cox:1964} is often applied, e.g., using
cubic \citep{Stidd:1973} or square root \citep{Hutchinson:1998} transformations.
However, the power parameter may vary for different climatic zones or temporal 
aggregation periods and hence we follow \cite{Stauffer+Mayr+Messner:2017} 
in their choice of $1.6^{-1}$ as a suitable power parameter for precipitation
in the region of Tyrol. The same power transformation is applied to both the
observed precipitation sums and the NWP outputs ``total precipitation'' (\emph{tp}) and
``convective available potential energy'' (\emph{cape}).

@@

\item \emph{Distributional forest:} All 80~predictor variables are
  considered for learning a forest of 100 trees. Subsampling is employed 
  for each tree using a third of the predictors
  in each split of the tree (\code{mtry}). Parameters are estimated by
  adaptive local likelihood based on the forest weights as described
  in Section~\ref{sec:methods}. 
  \hlc{The stopping criteria are the minimal number of observations to perform a split 
  (\texttt{minsplit} $= 50$), the minimal number of observations in a 
  segment (\texttt{minbucket} $= 20$), and the significance level 
  for variable selection (\texttt{alpha} $= 1$). The latter means
  that no early stopping (or ``pre-pruning'') is applied
  based on results of the statistical tests.}
  % stopping criteria:
  % forest_minsplit = 50, forest_minbucket = 20, forest_mincrit = 0

@@

In a first step, we show a detailed comparison of the competing models
for one observation site, Axams in Tyrol (in the center of the study area,
see Figure~\ref{fig:all-map}). As for all other stations, daily precipitation observations and
numerical weather predictions are available for the month of July from 1985
through 2012. In Figure~\ref{fig:axams-dist} in the introduction the probabilistic
forecasts from the distributional forest, trained on 1985--2008, for July 24
in 2009--2012 have already been shown as a motivational example. The figure
shows that the model properly
depicts the point masses at zero (i.e., the probability of a dry day) and
the forecasted probability density function for the total amount of precipitation.
The four sample forecasts differ considerably in location $\mu$, scale $\sigma$, and the amount of
censoring while conforming quite well with the actual observations
from these days. While this is a nice illustrative example we are  
interested in the overall predictive performance and calibration of the
distributional fits. \hlc{More details of this assessment as well as
an application to 14~further meteorological stations is provided in
Supplement~B} \citep{Schlosser+Hothorn+Stauffer:2018b}.

To assess calibration, Figure~\ref{fig:axams-qq} shows residual QQ~plots for
out-of-sample predictions (2009--2012) from the different models trained on
1985--2008. Due to the point masses at zero 100 draws from the randomized
quantile residuals \citep{Dunn+Smyth:1996} are plotted in semi-transparent
gray. Overall, the randomized quantile residuals conform quite well with the
theoretical standard normal quantile (i.e., form a straight line close to the 
diagonal), indicating that all four models are sufficiently well calibrated. 
\hlc{This is also supported by the corresponding probability integral transform}
(PIT, \citealp{Gneiting+Balabdaoui+Raftery:2007}) \hlc{histograms for station
Axams in Supplement B} \citep{Schlosser+Hothorn+Stauffer:2018b} \hlc{which 
contains a more detailed explanation of residual QQ plots and PIT histograms.}
